---
title: "Introduction to R"
subtitle: "Lecture 3: Data Visualization"
author: "Swapnil Singh"
date: "Lietuvos Bankas | [Course Link](https://github.com/swapnil1987/lb-introduction-to-r)" 
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{css, echo=FALSE}

@media print {
  .has-continuation {
    display: block !important;
  }
}

```

```{css, echo=FALSE}

.center-fit {
  display: block;
  margin-left: auto;
  margin-right: auto;
  max-width: 80%;
  max-height: 10%;
}


```


```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.


options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
   fig.asp = 9/16,
  fig.align = 'center',
  echo = F,
  out.width = "95%",
  dpi= 300
  )

knitr::opts_chunk$set(echo=TRUE, eval=TRUE,results='markup', cache = TRUE, 
                      warning = FALSE, message = FALSE)

Sys.setenv(lang='en')
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
library(tidyverse)
library(hrbrthemes)
library(fontawesome)
library(tidyverse)
library(ggtext)
library(lubridate)
library(here)
library(magick)
library(gapminder)
library(viridis)
library(palmerpenguins)
library(ggExtra)
library(janitor)
library(DBI)
library(dbplyr)
library(duckdb)


```



---

class: inverse, center, middle
name: bubble

# Basic Visualization

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>


---


# Why we need visualization?

- Imagine how different painters have drawn trees:
    - [Picasso](https://en.wikipedia.org/wiki/File:Pablo_Picasso,_1907-08,_Two_Trees_%28Les_Arbres%29,_watercolor_on_paper,_47.9_x_62.7_cm,_Philadelphia_Museum_of_Art.jpg)
    - [Monet](https://eclecticlight.co/2015/12/03/trees-in-the-landscape-8-claude-monet-and-his-poplar-series/)
    - [Mondrian](https://www.tate.org.uk/art/artworks/mondrian-the-tree-a-t02211)
    - [Kasiulis](https://artsandculture.google.com/asset/trees-by-the-river-vytautas-kasiulis-1918%E2%80%931995/ZwHhMuE3blBnCA?hl=en)
    
- Paintings convey
  - information
  - perspective of a painter

- Same is with data visualization
  - convey information in a concise manner
  - provide a perspective to look at the data
  

---

# What we will do in this class?

- Basics, nitty-gritty details of visualization
    - understand which type  of plot to use for different questions
    - understand different concepts within each plot
- Learn how `ggplot` helps us to visualize and convey information
    - create **presentable** graphics 
    

---

# Distributions

- When we want to understand only one variable, we look at the distribution
- Variables can be of two types:
    - discrete or categorical (take only few values)
        - Ex: gender, number of children in the household, ...
    - numerical or continuous (take many values)
        - Ex: age, population of different cities, GDP growth rate,...
        
---

# Distribution: Discrete Variable

- To visualize the distribution of a discrete variable, use `bar` chart
- For instance, in `penguins` data, we might be interested in the distribution of 
  penguin species

```{r, echo = TRUE, eval = FALSE}
ggplot(penguins, aes(x= species)) +
  geom_bar()

```

---

# Distribution: Discrete Variable

- To visualize the distribution of a discrete variable, use `bar` chart
- For instance, in `penguins` data, we might be interested in the distribution of 
  penguin species

```{r, echo = FALSE, eval = TRUE}
ggplot(penguins, aes(x= species)) +
  geom_bar()

```

---

# Distribution: Continuous Variable

- For continuous variables we can use `histogram`

```{r, echo=TRUE, eval=TRUE}

ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth=10)

```

---

# Distribution: Continuous Variable

- For continuous variables we can use `histogram`

```{r, echo=TRUE, eval=TRUE}

ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth=100)

```


---

# Distribution: Continuous Variable

- You can also use `density` which is just a smoothed histogram

```{r, echo=TRUE, eval=TRUE}

ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()

```

---

# Relationship between two variables

- Two variables can be 
    - continuous and discrete
    - discrete and discrete
    - continuous and continuous

---

# Continuous and Discrete

- Distribution of body mass by species of penguins

```{r}

ggplot(data = penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()

```


---

# Continuous and Discrete
- Distribution of body mass by species of penguins
```{r}

ggplot(data = penguins, aes( x = body_mass_g, color=species)) +
  geom_density()

```


---

# Two Discrete Variables

- Distribution of species across island: stacked bar plot
```{r}
ggplot(data = penguins, aes(x = island, fill=species)) +
  geom_bar()

```


---

# Two Discrete Variables

- Distribution of species across island: side by side bar plot

```{r}
ggplot(data = penguins, aes(x = island, fill=species)) +
  geom_bar(position = 'dodge')

```


---

# Two continuous variables

- From the last class

```{r, echo = FALSE, cache=TRUE, warning=FALSE, message=FALSE, eval=TRUE}
p <-  ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  
  
  # notice how `aes` for geom_point shifted from above to here
  geom_point(mapping = aes(color=species, shape=species)) +
  
  
  #add linear fit (one line across all three groups)
  geom_smooth(method='lm')  +
  
  #modern theme
  theme_modern_rc() +
  
  labs(
    title = "Penguins: relationship between body mass and flipper length",
    subtitle = "Adelie, Chinstrap, and Gentoo Penguins",
    x = "flipper length in millimeters",
    y = "body mass in grams",
    color = "species",
    shape = "species"
  )

p
```


---

# Two continuous variables

- From the last class (conditional on island)

```{r, echo = TRUE, cache=TRUE, warning=FALSE, message=FALSE, eval=FALSE}
p <-  ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  
  
  # notice how `aes` for geom_point shifted from above to here
  geom_point(mapping = aes(color=species, shape=species)) +
  
  
  #add linear fit (one line across all three groups)
  geom_smooth(method='lm')  +
  
  #conditional on island
  facet_wrap(~island) +
  
  #modern theme
  theme_modern_rc() +
  
  labs(
    title = "Penguins: relationship between body mass and flipper length",
    subtitle = "Adelie, Chinstrap, and Gentoo Penguins",
    x = "flipper length in millimeters",
    y = "body mass in grams",
    color = "species",
    shape = "species"
  )

p
```

---

# Two continuous variables

- From the last class (conditional on island)

```{r, echo = FALSE, cache=TRUE, warning=FALSE, message=FALSE, eval=TRUE}
p <-  ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  
  
  # notice how `aes` for geom_point shifted from above to here
  geom_point(mapping = aes(color=species, shape=species)) +
  
  
  #add linear fit (one line across all three groups)
  geom_smooth(method='lm')  +
  
  #conditional on island
  facet_wrap(~island) +
  
  #modern theme
  theme_modern_rc() +
  
  labs(
    title = "Penguins: relationship between body mass and flipper length",
    subtitle = "Adelie, Chinstrap, and Gentoo Penguins",
    x = "flipper length in millimeters",
    y = "body mass in grams",
    color = "species",
    shape = "species"
  )

p
```


---

class: inverse, center, middle
name: marginal

# Marginal Distributions

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Marginal Distributions

```{r, echo=FALSE}

p <- ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g)) +
      
  # notice how `aes` for geom_point shifted from above to here
  geom_point(mapping = aes(color=species, shape=species)) +
  
  
  #add linear fit (one line across all three groups)
  geom_smooth(method='lm')  +

  #modern theme
  theme_bw() +
  
  labs(
    title = "Relationship between body mass and flipper length",
    subtitle = "Adelie, Chinstrap, and Gentoo Penguins",
    x = "flipper length in millimeters",
    y = "body mass in grams",
    color = "species",
    shape = "species"
  ) +
  theme(legend.position = 'none')

 
# with marginal histogram
p1 <- ggMarginal(p, type="histogram")
p1

```


---


# Marginal Distributions

```{r, echo=TRUE, eval=FALSE}

p <- ggplot(penguins, aes(x=flipper_length_mm, y=body_mass_g)) +
      
  # notice how `aes` for geom_point shifted from above to here
  geom_point(mapping = aes(color=species, shape=species)) +
  
  
  #add linear fit (one line across all three groups)
  geom_smooth(method='lm')  +

  #modern theme
  theme_bw() +
  
  labs(
    title = "Relationship between body mass and flipper length",
    subtitle = "Adelie, Chinstrap, and Gentoo Penguins",
    x = "flipper length in millimeters",
    y = "body mass in grams",
    color = "species",
    shape = "species"
  ) +
  
  theme(legend.position = 'none')

```

---

# Marginal Distributions


```{r, echo=TRUE, eval=TRUE}

# with marginal histogram
p1 <- ggMarginal(p, type="histogram")
p1


```

---

# Marginal Distributions

```{r}

# with marginal histogram
p2 <- ggMarginal(p, type="histogram", fill='red')
p2

```


---

class: inverse, center, middle
name: data-import

# Importing  Data in R

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html>

---

# Reading csv files

- Comma separated value (CSV) files are most common
- Command is: `read_csv`

```{r}
cleanFuelData <- read_csv(file = 'raw_data/clean-fuel-data.csv')
glimpse(cleanFuelData)
```

---

# Reading csv files

- How to make sure that `..` is recognized as missing

```{r}
cleanFuelData <- read_csv(file = 'raw_data/clean-fuel-data.csv',
                          na = c('..'))
glimpse(cleanFuelData)
```

---

# Reading csv files
- Now look at variable names like country name or country code
    - surrounded by backticks. Why?
        - space between names
- we need to clean up in a cumbersome way

```{r}
cleanFuelData <- read_csv(file = 'raw_data/clean-fuel-data.csv',
                          na = c('..')) |> 
  rename(country_name = `Country Name`,
         country_code = `Country Code`)

```

---

# Reading csv files
- Alternative way to clean variable names: use `janitor` package
```{r}
cleanFuelData <- read_csv(file = 'raw_data/clean-fuel-data.csv',
                          na = c('..')) |> 
  janitor::clean_names()
glimpse(cleanFuelData)
```


---

# Importing from Databases

- CSV or Excel are not always available
- Sometimes, you have to access `databases`
- We need to learn two things then
    - `DBI` package
    - `dbplyr` package
- `DBI` package connects you to a database
- `dbplyr` converts `dplyr` code to `SQL` 


---

# What is a database?

- Think of database as a bunch of `dataframes`
- These `dataframes` are also called `tables`
    - there are some difference between `dataframe` and `table` but not important for this class
- Databases are run by database management systems (DBMS)
- Three types of DBMS
    1. Client server
    2. Cloud
    3. In-process


---

# How it works? Big Picture

- There are two steps involved:
    1. use `DBI` to connect to the database and perform simple functions
    2. Depending on the DBMS, you will need specific package
        - `RPostgres` for `PostgreSQL`
        - `RMariaDB` for `MySQL`
- For this class we will use in-house `duckdb`
    - difference between `duckdb` and other DBMS is only how you connect to it
    - everything else is essentially the same
    
---

# Working with duckdb

```{r}
#create empty database
con <- dbConnect(duckdb(), dbdir='chapter3_db')

# add some data to it
dbWriteTable(con, "penguins", palmerpenguins::penguins, overwrite=TRUE)
dbWriteTable(con, "penguins_raw", palmerpenguins::penguins_raw, overwrite=TRUE)
dbWriteTable(con, "diamonds", ggplot2::diamonds, overwrite=TRUE)

#now check which tables are in the database
dbListTables(con)

# pull one of the tables
con |> dbReadTable('penguins') 
```


---

# Introducing dbplyr

```{r}
con <- dbConnect(duckdb(), dbdir='chapter3_db')

# add some data to it
dbWriteTable(con, "penguins", palmerpenguins::penguins, overwrite=TRUE)
dbWriteTable(con, "penguins_raw", palmerpenguins::penguins_raw, overwrite=TRUE)
dbWriteTable(con, "diamonds", ggplot2::diamonds, overwrite=TRUE)

penguins_db <- tbl(con, 'penguins')
penguins_db
```




```{r gen_pdf, include = FALSE, cache = FALSE, eval = TRUE}
infile = list.files(pattern = '.html')
pagedown::chrome_print(input = infile, timeout = 100)
```


