---
title: "Introduction to R"
subtitle: "Lecture 2: Introduction to `tidy` Environment"
author: "Swapnil Singh"
date: "Lietuvos Bankas | [Course Link](https://github.com/swapnil1987/lb-introduction-to-r)" 
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{css, echo=FALSE}
# CSS for including pauses in printed PDF output (see bottom of lecture)
@media print {
  .has-continuation {
    display: block !important;
  }
}
```

```{css, echo=FALSE}
.center-fit {
  display: block;
  margin-left: auto;
  margin-right: auto;
  max-width: 80%;
  max-height: 10%;
}
```


```{r setup, include=FALSE}
# xaringanExtra::use_scribble() ## Draw on slides. Requires dev version of xaringanExtra.


options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
   fig.asp = 9/16,
  fig.align = 'center',
  echo = F,
  out.width = "95%",
  dpi= 300
  )

knitr::opts_chunk$set(echo=TRUE, eval=TRUE,results='hide')

Sys.setenv(lang='en')
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
library(tidyverse)
library(hrbrthemes)
library(fontawesome)
library(tidyverse)
library(ggtext)
library(lubridate)
library(here)
library(magick)
library(gapminder)
```



# Lecture's Objectives

1. Quick overview of how to transform the data
    -  operation on `rows`
    -  operation on `columns`
    -  operation `==` verbs
2. Learn about `pipes`
    -  Concatenation of verbs through pipes
3. How to work with `groups`

---

# Illustrative Data

- We are going to use the `gapminder` data

```{r, cache=TRUE, include=TRUE, results='markup'}

head(gapminder)
colnames(gapminder)
nrow(gapminder)

```

---

# Illustrative Data

- We are going to use the `gapminder` data

```{r, cache=TRUE, include=TRUE, results='markup'}
glimpse(gapminder)
```

- Notice the difference between `head` and `glimpse`

---

# Difference: _tibble_ and _dataframe_

- `tibbles` are special types  of data frame
- Main differences
    1. tibbles are designed for large datasets
    2. tibble prints are concise: only few rows and only those columns that fit the screen
- To view the full dataset:
    - `View(gapminder)`: interactive dataviewer
    - `print(gapminder, width = Inf)`: show all  columns
- **Recommendation**
    -  Use the combination of `glimpse()` and `View()` functions
    


---

# Data transformation using `dplyr`

- To transform data, you have to rely on `dplyr` package's functions
- Automatically loads  with command `library(tidyverse)`
- Huge repository of functions
```{r, cache=TRUE, include=TRUE, results='markup'}

#number of functions in dplyr package
funcsDplyr <- ls('package:dplyr')  #note the package has to be loaded first
numFuncsDplyr  <- length(funcsDplyr)
numFuncsDplyr

#first twenty functions
print(funcsDplyr[1:20])

```

---

# Data transformation using `dplyr`
    
- Each function in `dplyr` perform a **single** task
- Data transformation requires multiple task
- Single task $\rightarrow$ multiple task: use `pipes` i.e. `|>`
- An inscrutable example at this moment

```{r, echo=TRUE, eval=TRUE, cache=TRUE, results='markup'}
newGapminder <- gapminder |> 
  
  # keep only Asian continent
  filter(continent == "Asia") |> 
  
  # life  expectancy > 70
  filter(lifeExp > 70)
```

--

```{r, echo=TRUE, eval=TRUE, cache=TRUE, results='markup'}
glimpse(newGapminder)
```

---

# Organization of dplyr functions

-  Functions can be broadly classified as
    - `row`  functions
    - `column` functions
    - `group` functions
    - `tables` functions
-  This course we mostly cover the first three
- Let's  deal them one by one


---

# Row functions: _filter_

- `filter`
    - keep only rows satisfying certain criteria
    
```{r, echo=TRUE, eval=FALSE, results='markup', cache=TRUE}

#keep only those observation with gdp per capital greater thann 1000
newGapminder <-  gapminder |> 
  filter(lifeExp > 1000)
```

- Operations
    - _>_ greater than
    - _>=_ greater than or equal to
    - < less than
    - <=  less than or equal  to
    - == equal to
    - != not equal to

- Combination: 
    - `&` and
    - `|` or

---


# Row functions: _filter_

```{r, echo=TRUE, eval=TRUE, results='markup'}

 gapminder |> 
  
  # focus only on Asian countries with life expectancy greater than 70
  filter(lifeExp > 70 & continent=='Asia') |> 
  
  #focus  only on countries with per capita GDP greater than 2000
  filter(gdpPercap > 2000)
  
  

```


---

# Row functions: _filter_

- Use `%in%` when combining `|` and `==`

```{r, cache=TRUE, results='markup'}
gapminder |> 
  filter(country == 'Afghanistan' | country == 'Mali' )

```



---

# Row functions: _filter_

```{r, cache=TRUE, results='markup'}
gapminder |> 
  filter(country %in% c('Afghanistan', 'Mali') )

```


---

# Row functions: _filter_

- When filtering, `dplyr::filter` operates on `gapminder` and prints the final results
- `gapminder` data by itself is untouched
- You can store the result in the new `dataframe`

```{r, cache=TRUE, results='markup'}
newGapminder <-  gapminder |> 
  
  filter(country %in% c('Afghanistan', 'Mali') )

newGapminder
```


---


# Row functions: _arrange_

- Change  the ordering of rows based on the value of column(s)

```{r, cache=TRUE, results='markup'}
gapminder |> 
  arrange(year, gdpPercap) 
```

---


# Row functions: _arrange_


- Use `desc` within `arrange` for sorting from largest to smallest
    - Caveat: can use only one argument

```{r, cache=TRUE, results='markup'}
gapminder |> 
  arrange(desc( gdpPercap)) 
```


---

# Row functions: _distinct_

- _distinct_ searches and keeps all unique  rows in a dataset

```{r, cache=TRUE, results='markup'}

gapminder |> 
  distinct()
```


---

# Row functions: _distinct_

- _distinct_ on it own is not very useful
- When we want to keep _distinct_ combination of some variables, it is very useful


```{r, cache=TRUE, results='markup'}

gapminder |> 
  distinct(year) |> 
  head(n=4)
```


---

# Row functions: _distinct_


- Note that only `year` column is kept
- To rectify, use `.keep_all = TRUE`


```{r, cache=TRUE, results='markup'}

gapminder |> 
  distinct(year, .keep_all = TRUE) |> 
  head(n=4)
```


---

# Column functions: _mutate_
- _mutate_ helps add new columns using existing columns

```{r, cache=TRUE, results='markup'}
gapminder |> 
  mutate(logGdpPerCapita = log(gdpPercap))

```



---

# Column functions: _mutate_

- Imagine you have a dataset of `500` columns
- You do a `mutate` and then `View` it
- By default new column is added  to the last
- Use `.before` to put the new column first
```{r, cache=TRUE, results='markup'}
gapminder |> 
  mutate(logGdpPerCapita = log(gdpPercap), .before=1)

```

---

# Column functions: _mutate_

-  Can use `.after` also
```{r, cache=TRUE, results='markup'}
gapminder |> 
  mutate(logGdpPerCapita = log(gdpPercap), .after=year)

```

---

# Column functions: _select_




